Epistemic Semantics in Guarded String Models

### 2020-06-18 tex/draft.tex
polish 5

\section{Direct Bounded Interpretation}

@@

We translate the Epik terms into a Haskell algebraic datatype that
represents terms with the same signature as described in Section 2. We
then paramaterize the interpretation function, $\llparenthesis p
\rrparenthesis^{\varphi,{\mathcal{E}}}_n$ on an integer $n$ that
describes the maximum length of a string we will produce. The bounded
interpretation into lists of strings is very similar to the unbounded
interpretation into sets of strings, except for the bounds
checking. The full details are shown in Figure~\ref{fig:list-interp}:
we use list-comprehension notation, which is analogous to set-builder
notation, except that it produces a list (potentially with duplicated
elements) instead of a set, and so is written with square brackets
$[]$. We use $+$ to indicate list concatenation, and $\sum l$ to
indicate the concatenation of a list of lists $l$. We overload the set
membership symbol $\in$ to also indicate membership in a list. We also
lift the fusion operator $\diamond$ to lists. For a list of guarded
strings $l$, we write $l|_n$ as shorthand for the set $[ g~|~g \in l ,
  |g| \leq n]$. Finally, we write $\lfloor X \rfloor$ to convert a set
$X$ into a list containing same elements with an arbitrary order.

First of all, whenever $n = 0$, the set of guarded strings is empty. TODO ADD
MORE TEXT.

@@

\begin{figure}
  \[\begin{array}{r >{\triangleq}c l}
  \llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_0 && [] \\
  \llparenthesis 0 \rrparenthesis^{\varphi,{\mathcal{E}}}_n && [] \\
  \llparenthesis 1 \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \lfloor \mathcal A_{\mathsf B}^\varphi \rfloor \\
  \llparenthesis e \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \lfloor \hat{\mathcal E}^\varphi(e)\rfloor \\
  \llparenthesis \psi \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \lfloor \mathcal A_{\mathsf B}^{\varphi~\psi} \rfloor \\
  \llparenthesis p + q \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_n + \llparenthesis q \rrparenthesis^{\varphi,{\mathcal{E}}}_n \\
  \llparenthesis p;q \rrparenthesis^{\varphi,{\mathcal{E}}}_n &&
  \left(\llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_n \cdot \llparenthesis q\rrparenthesis^{\varphi,{\mathcal{E}}}_n\right)|_n\\
  \llparenthesis p^* \rrparenthesis^{\varphi,{\mathcal{E}}}_n && [] + \left(\llparenthesis p\rrparenthesis^{\varphi,{\mathcal{E}}}_n \diamond  \llparenthesis p^* \rrparenthesis^{\varphi,{\mathcal{E}}}_{n-i}\right)|_n \\
  \multicolumn 2 r {} & \text{where}~i = \max\{1,\min\{|g|\mid g \in \llbracket p \rrbracket^{\varphi,{\mathcal{E}}}\}\}\\
  \llparenthesis \neg p \rrparenthesis^{\varphi,{\mathcal{E}}}_n && [g \mid g \in \llbracket \_^* \rrbracket^{\varphi,{\mathcal{E}}}_n, g \not\in \llbracket p \rrbracket^{\varphi,{\mathcal{E}}}_n] \\
  \llparenthesis \Diamond_a~p \rrparenthesis^{\varphi,{\mathcal{E}}}_n &&
  [g' \mid g \in g'~\hat{R_a}~g, g \in \llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}]
  \end{array}
  \]
  \caption{Bounded interpretation using lazy lists}
\end{figure}

TODO Describe advantages of Lazy Lists

>>>>

\section{Bounded Lazy Interpretation}

@@

We translate the Epik terms into a Haskell algebraic datatype that
represents terms with the same signature as described in Section 2. We
then paramaterize the interpretation function, $\llparenthesis p
\rrparenthesis^{\varphi,{\mathcal{E}}}_n$ on an integer $n$ that
describes the maximum length of a string we will produce.

The bounded interpretation into lists of strings is very similar to
the unbounded interpretation into sets of strings, except for the
bounds checking. The full details are shown in
Figure~\ref{fig:list-interp}. First note that when $n=0$, we simply
return the empty list, denoted $[]$. Terms of the form $0$, $1$, $e$,
and $\psi$ have the same denotation as before, translated into a list
(for a set $X$, $\lfloor X \rfloor$ is a list with the same elements
as $X$ arbitrarily ordered). The term $p+q$ denotes the list
concatenation (written $\texttt{+\!+}$) of the denotation of $p$ and of
$q$. The term $p;q$ denotes the fusion product (lifted to lists) of
the denotations of $p$ and in $q$ restricted to only those strings
shorter than $n$ (for a list $l$, $l|_n$ filters out elements longer
than $n$). The denotation of $p^*$ uses the fact that $p^*$ and $1 +
p;p^*$ are equivalent, and decrements the size threshold on the
recursive denotation of $p^*$ by $i$, where $i$ is the length of the
longest (nonzero) string in the denotation of $p$, making sure to
filter out guarded strings that are too long. The denotation of $\neg
p$ is the strings that occur in $\_^*$ and not in $p$ (the $\setminus$
operator on lists is analogous to the $\setminus$ operator on
sets). The denotation of $\Diamond_a~p$ is analogous to the set
denotation, depicted in Figure~\ref{fig:list-interp}-comprehension
notation\footnote{List comprehension notation is analogous to set
  builder notation, except that it is written using square
  brackets. The order preservation is indicated by the keyword
  \texttt{for}}.

@@

\begin{figure}
  \[\begin{array}{r >{\triangleq}c l}
  \llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_0 && [] \\
  \llparenthesis 0 \rrparenthesis^{\varphi,{\mathcal{E}}}_n && [] \\
  \llparenthesis 1 \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \lfloor \mathcal A_{\mathsf B}^\varphi \rfloor \\
  \llparenthesis e \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \lfloor \hat{\mathcal E}^\varphi(e)\rfloor \\
  \llparenthesis \psi \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \lfloor \mathcal A_{\mathsf B}^{\varphi~\psi} \rfloor \\
  \llparenthesis p + q \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_n \mathbin{\texttt{+\!+}} \llparenthesis q \rrparenthesis^{\varphi,{\mathcal{E}}}_n \\
  \llparenthesis p;q \rrparenthesis^{\varphi,{\mathcal{E}}}_n &&
  \left(\llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_n \diamond \llparenthesis q\rrparenthesis^{\varphi,{\mathcal{E}}}_n\right)|_n\\
  \llparenthesis p^* \rrparenthesis^{\varphi,{\mathcal{E}}}_n && [] + \left(\llparenthesis p\rrparenthesis^{\varphi,{\mathcal{E}}}_n \diamond  \llparenthesis p^* \rrparenthesis^{\varphi,{\mathcal{E}}}_{m}\right)|_n \\
  \multicolumn 3 r {\text{where}~
    \begin{array}[t]{r >{\hspace{-.7em}}l}
      i &= \max\{1,\min\{|g|\mid g \in \llparenthesis p \rrparenthesis_n^{\varphi,{\mathcal{E}}}\}\}\\
      m &= \max\{0, n - i\}
    \end{array}}\\
  \llparenthesis \neg p \rrparenthesis^{\varphi,{\mathcal{E}}}_n && \llparenthesis \_^* \rrparenthesis^{\varphi,{\mathcal{E}}}_n \setminus \llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_n \\
  \llparenthesis \Diamond_a~p \rrparenthesis^{\varphi,{\mathcal{E}}}_n &&
  [g' \mid g'~\hat{R_a}~g,~\texttt{for}~g~\texttt{in}~\llparenthesis p \rrparenthesis^{\varphi,{\mathcal{E}}}_n]
  \end{array}
  \]
  \caption{Bounded interpretation using lazy lists}
  \label{fig:bound-interp}
\end{figure}

The lists we use here are \emph{lazy} (as opposed to \emph{strict}),
which broadly means that computation is delayed until the value is
needed. This allows us to avoid computing large, unnecessary
iterations. In the elevator example, this means that $\hat 0;u;u;u^*$
will unroll the $u^*$ 0 times, once, and then discover that it is
impossible to unroll it a second time, because $\llparenthesis\hat
3;u\rrparenthesis^{\varphi,\mathcal E}_n = []$, for any $n$. So
$\llparenthesis\hat 3;u\rrparenthesis^{\varphi,\mathcal E}_n \diamond
\llparenthesis u^*\rrparenthesis^{\varphi,\mathcal E}_{n-1}$ will
return $[]$ without having to compute the full fixpoint of
$\llparenthesis u\rrparenthesis^{\varphi,\mathcal E}$. Similar
behavior occurs once $n = 0$.

Conversely, if we used sets (as in the math) instead of lists we would
need to constantly verify the set invariant (that every element is
unique) which means processing every element in the set. Verifying
that every element in $\llparenthesis
u^*\rrparenthesis^{\varphi,\mathcal E}$ is unique would require the full
computation of the fixpoint.
